A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-01-23 paper/paper.tex
paper: fix line breaks before align*
paper: clip whitespace around equation*

Our precise mechanization in Coq is available at \href{https://github.com/artagnon/bonak}{github.com/artagnon/bonak}. The construction was conceived in Summer 2019 and the mechanization began in late 2019. A sketch of the construction was presented at the 2020 HoTT-UF workshop and the completion of the mechanization was reported at the TYPES 2022 conference.

\section{Semi-simplicial and semi-cubical sets}
It is well-known that simplicial sets exist purely in the imagination of the mathematician, as every simplicial set has an infinite number of degenerate simplices. What we have formalized is \emph{augmented semi-simplicial} sets~\cite{fri08}, and \emph{semi-cubical} sets~\cite{grandis03,buchholtz17}. We will proceed to explain the mathematical objects we have formalized. % TODO: well-known: can we find a reference?

@@

\begin{definition}[$\Set_{\DeltaPlus}$]
  We define the category of augmented semi-simplicial sets as the functor category:

  \begin{equation*}
    \Set_{\DeltaPlus} := \Set^{\DeltaPlus^{\mathsf{op}}}
  \end{equation*}
\end{definition}

@@

\begin{definition}[$\DeltaPlus^n$]
  The standard augmented $n$-semi-simplex $\DeltaPlus^n$ is defined as what is called the Yoneda embedding of $n \in \Obj(\DeltaPlus)$:

  \begin{align*}
    \DeltaPlus^n    & : \Set_{\DeltaPlus}        \\
    \DeltaPlus^n(p) & := \Hom(p, n)              \\
    \DeltaPlus^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

\begin{example}[$\DeltaPlus^1$]
  In dimension $1$, the standard augmented semi-simplex can be pictured as a point, colored black, corresponding to the unique morphism in $\Hom(0, 1)$. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$:

  \begin{equation*}
    \begin{tikzcd}
      \kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^2$]
  In dimension $2$, the standard augmented semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use the color black to denote the unique morphisms in $\Hom(0, 1)$ and $\Hom(0, 2)$.

  \begin{equation*}
    \begin{tikzcd}
      \kstar0 \arrow[r, dash, "\kstar\kstar"] & 0\kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^3$]
  $\DeltaPlus^3$ is drawn as three points, given by $\Hom(1, 3)$, three lines connecting them, given by $\Hom(2, 3)$, and a triangular filler given by $\Hom(3, 3)$.

  \begin{equation*}
    \begin{tikzcd}
      & |[alias=F]|00\kstar \arrow[ddr, dash, "0\kstar\kstar"] & \\\\
      \kstar00 \arrow[rr, dash, "\kstar\kstar0"{name=T, below}]\arrow[uur, dash, "\kstar0\kstar"] && 0\kstar0 \\
      \arrow[rightarrow, from=F, to=T, phantom, "\kstar\kstar\kstar" description]
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\begin{definition}[\CSet]
  We define the category of semi-cubical sets as the functor category:

  \begin{equation*}
    \CSet := \Set^{\op{\Cube}}
  \end{equation*}
\end{definition}

\begin{definition}[$\Cube^n$]
  The standard semi-cube $\Cube^n$ is defined as the Yoneda embedding of $n \in \Obj(\Cube)$:

  \begin{align*}
    \Cube^n    & : \CSet                    \\
    \Cube^n(p) & := \Hom(p, n)              \\
    \Cube^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

\begin{example}[$\Cube^0$]
  $\Cube^0$ is $\Hom(0, 0)$, or the singleton set of the empty sequence:

  \begin{equation*}
    \begin{tikzcd}
      \{\}
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\Cube^1$]
  $\Cube^1$ consists of two points, given by $\Hom(0, 1)$ and a line, given by $\Hom(1, 1)$:

  \begin{equation*}
    \begin{tikzcd}
      L \arrow[r, "\kstar", dash] & R
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\Cube^2$]
  $\Cube^2$ consists of four points, given by $\Hom(0, 2)$, four lines connecting the four points, given by $\Hom(1, 2)$, and a filler, given by $\Hom(2, 2)$:

  \begin{equation*}
    \begin{tikzcd}
      LR \arrow[r, dash, "\kstar R"{name=F}] \arrow[d, dash, "L\kstar" left] & RR \arrow[d, dash, "R\kstar"] \\
      LL \arrow[r, dash, "\kstar L"{name=T, below}] & RL \\
      \arrow[rightarrow, from=F, to=T, phantom, "\kstar\kstar" description]
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\nu$-semi-shape category. The morphisms of the $\nu$-semi-shape category are sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$, so that the following holds:

@@

\section{Our setting}
In this section, we explain how we recover the notion of sets in type theory, and briefly explain the difference between the fibered and indexed representations.

@@

A notion of sets can be recovered in each universe as $\U[m]$, denoting the subtype of $\Type_m$ for which paths are degenerated, what can be expressed by the property of Uniqueness of Identity Proofs (UIP). Technically, this is expressed as a structure equipping a domain $\Dom$ with the property $\UIP$:

\begin{align*}
  \Dom & : \Type_m                                  \\
  \UIP & : \Pi x y: \Dom.\, \Pi p q: x = y.\, p = q
\end{align*}

@@

Note that an equivalent definition of UIP is given by \textsf{UIP\_refl}:

\begin{align*}
  \textsf{UIP\_refl} & : \Pi x: \Dom.\, \Pi p: x = x.\, p = \textsf{eq\_refl } x
\end{align*}

@@

The latter correspondence can be expressed in Homotopy Type Theory, for $T: \Type$, as an equivalence between $\Sigma S: \Type\;.\;S \rightarrow T$ and $T \rightarrow \Type$, which can be proved using univalence. This correspondence also holds on \U, for $T: \U$:

\begin{equation*}
  \Sigma S: \U\;.\;S\rightarrow T ~\simeq~ T \rightarrow \U
\end{equation*}

\subsection{Example: indexed semi-cubical sets}

Let us now come back to semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.

\begin{equation*}
  \begin{tikzcd}
    X_0 : \U & X_1 : \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2 : \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions. Based on these coherence conditions, the type of $X_2$ happens to have the form $\Pi a b c d, X_1 (a)(b) \rightarrow X_1 (a)(c) \rightarrow X_1 (b)(d) \rightarrow X_1 (c)(d) \rightarrow \U$, recovering the square structure of $X_2$. In summary,

\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

@@

\subsection{Relating the indexed representation to parametricity}
The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:

\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

In particular, for $X: \U$, applying the parametricity translation is about canonically associating to $X$ an inhabitant $X_\kstar$ of ${\U}_\kstar(X)(X)$ i.e. of $X \rightarrow X \rightarrow \U$. In turn, applying the parametricity translation to $X_\kstar: X \rightarrow X \rightarrow \U$ is about canonically associating to $X_\kstar$ an inhabitant $X_{\kstar\kstar}$ of $(X \rightarrow X \rightarrow \U)_\kstar(X_\kstar)(X_\kstar)$ i.e. of:

\begin{align*}
  \Pi x_{LL}: X.\, \Pi x_{LR}: X.\, X_\kstar(x_{LL})(x_{LR})             \\
  \rightarrow \Pi x_{RL}: X.\, \Pi x_{RR}: X.\, X_\kstar(x_{RL})(x_{RR}) \\ \rightarrow X_\kstar(x_{LL}, x_{RL}) \rightarrow X_\kstar(x_{LR})(x_{RR})  \rightarrow \U
\end{align*}
which hints us at how the sequence $X_0$, $X_1$, $X_2$ can be seen as a sequence of inhabitants of the iteration of binary parametricity applied to an initial $X: \U$:

\begin{equation*}
  \begin{array}{llllr}
    X_0 & \defeq & X                & : & \U                                               \\
    X_1 & \defeq & X_\kstar         & : & {\U}_\kstar(X)(X)                                \\
    X_2 & \defeq & X_{\kstar\kstar} & : & ({\U}_\kstar(X)(X))_{\kstar}(X_\kstar)(X_\kstar) \\
    \ldots
  \end{array}
\end{equation*}

@@

\subsection{Intuition for our formal construction}
To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on building blocks, which we call $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube, etc.), which we decompose into $\layer$s, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$.

@@

We illustrated here the cubical case, that is $\nu = 2$, but, in general, a $\layer[n][p]$ is a product of $\nu$ $\painting[n - 1][p]$. A $\painting[n][0]$ is a $n$-dimensional object corresponding to a filled $\fullframe[n]$. More generally, a $\painting[n][p]$ is an $n$-dimensional object which has the form of a $\painting[n-p][0]$, thus of $(n-p)$-dimensional form, but shifted and living in dimensions $p$ to $n$. Such $\painting[n][p]$ fills a space framed by a partial $\framep[n][p]$ so that, together, they form a filled $\fullframe[n]$. For instance, in the picture, each of the two $\painting[2][0]$ of $\layer[3][0]$ is a filled blue square, each of the two $\painting[2][1]$ of $\layer[3][1]$ is the line shown in red, stretched into a partial square filling the partial frames made of respectively the left and right border of the blue square, and each of the two $\painting[2][2]$ of $\layer[3][2]$ is a point shown in green, stretched into a partial square filling the full frames made respectively of the upper and lower borders of the blue and red squares. A $\painting[n][p]$ adds to a $\framep[n][p]$ the layers needed to form a $\fullframe[n]$. They are added from dimension $n$ to dimension $p$ that is opposite to the order from $0$ to $p$ the $\framep[n][p]$ are built. A $\painting[n][p]$ complementing a given $\framep[n][p]$ fills the resulting $\fullframe[n]$. Thus, on the contrary of frames, it includes a filler, which lives in $X_n$ as shown below:

\begin{align*}
  \framep[n][p]   & \defeq \Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0]) \ldots) .\; \layer[n][p] \\
  \painting[n][p] & \defeq \Sigma l_p : \layer[n][p] . (\ldots (\Sigma l_n : \layer[n][n - 1] .\; X_n) \ldots)
\end{align*}

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a type, which we call $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then refine the previous equation by showing the dependencies on $D$. In particular, $X_n$ is just $D.\tl$.

\begin{align*}
   & \framep[n][p](D) \defeq                                                                                     \\
   & \;\;\;\;\Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0](D)) \ldots) .\,\layer[n][p](D) \\
   & \painting[n][p](D) \defeq                                                                                   \\
   & \;\;\;\;\Sigma l_p : \layer[n][p](D).\, (\ldots (\Sigma l_n : \layer[n][n - 1](D) .\, D.\tl) \ldots)
\end{align*}

@@

Let us recall that a layer is made of $\nu$ paintings. When $\nu = 2$, this amounts to:

\begin{align*}
  \layer[n][p](D)(d) & \defeq & \painting[n-1][p](D.\hd)(\restrf[n][p][L][p](d)) & \times \\
                     &        & \painting[n-1][p](D.\hd)(\restrf[n][p][R][p](d)) &
\end{align*}

@@

Given $c_\omega$ of type

\begin{align*}
  c_\omega & : \painting[n-1][p](D.\hd)(\restrf[n-1][p][\omega][q](d))
\end{align*}
$\restrl[n][p][\epsilon](d)(c_L, c_R)$ produces a layer, in which the $\omega$-component has the type

\begin{equation*}
  \painting[n-2][p](D.\hd.\hd)(\restrf[n-1][p][\epsilon][q](\restrf[n][p][\omega][p](d)))
\end{equation*}
while we expect a term of type

\begin{equation*}
  \painting[n-2][p](D.\hd.\hd)(\restrf[n-1][p][\omega][p](\restrf[n][p][\epsilon][q+1](d)))
\end{equation*}

@@

We now present the construction in \emph{extensional} type theory; i.e. in a type theory with the following reflection rule, where $=$ is propositional equality in some type and $\equiv$ is definitional equality~\cite{martinlof84}:

\begin{equation*}
  \seqr{}{\Gamma \vdash p : t = u}{\Gamma \vdash t \equiv u}
\end{equation*}

@@

where $\mathsf{ap}$ applies a function on two sides of an equality, and $\bullet$ is transitivity of equality. This property of equality proofs holds in \U, and since our formalization is done in \U, the term is trivially discharged.

>>>>

Our precise mechanization in Coq is available at \href{https://github.com/artagnon/bonak}{github.com/artagnon/bonak}. The construction was conceived in Summer 2019, and the mechanization began in late 2019. A sketch of the construction was presented at the 2020 HoTT-UF workshop and the completion of the mechanization was reported at the TYPES 2022 conference.

\section{Semi-simplicial and semi-cubical sets}
It is well-known that simplicial sets exist purely in the imagination of the mathematician, as every simplicial set has an infinite number of degenerate simplices. What we have formalized is \emph{augmented semi-simplicial} sets~\cite{fri08}, and \emph{semi-cubical} sets~\cite{grandis03,buchholtz17}. We will proceed to explain the mathematical objects we have formalized, the connection with parametricity, and our generalization to $\nu$-sets. % TODO: well-known: can we find a reference?

@@

\begin{definition}[$\Set_{\DeltaPlus}$]
  We define the category of augmented semi-simplicial sets as the functor category:
  \begin{equation*}
    \Set_{\DeltaPlus} := \Set^{\DeltaPlus^{\mathsf{op}}}
  \end{equation*}
\end{definition}

@@

\begin{definition}[$\DeltaPlus^n$]
  The standard augmented $n$-semi-simplex $\DeltaPlus^n$ is defined as what is called the Yoneda embedding of $n \in \Obj(\DeltaPlus)$:
  \begin{align*}
    \DeltaPlus^n    & : \Set_{\DeltaPlus}        \\
    \DeltaPlus^n(p) & := \Hom(p, n)              \\
    \DeltaPlus^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

\begin{example}[$\DeltaPlus^1$]
  In dimension $1$, the standard augmented semi-simplex can be pictured as a point, colored black, corresponding to the unique morphism in $\Hom(0, 1)$. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$:
  \begin{equation*}
    \begin{tikzcd}
      \kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^2$]
  In dimension $2$, the standard augmented semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use the color black to denote the unique morphisms in $\Hom(0, 1)$ and $\Hom(0, 2)$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar0 \arrow[r, dash, "\kstar\kstar"] & 0\kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^3$]
  $\DeltaPlus^3$ is drawn as three points, given by $\Hom(1, 3)$, three lines connecting them, given by $\Hom(2, 3)$, and a triangular filler given by $\Hom(3, 3)$.
  \begin{equation*}
    \begin{tikzcd}
      & |[alias=F]|00\kstar \arrow[ddr, dash, "0\kstar\kstar"] & \\\\
      \kstar00 \arrow[rr, dash, "\kstar\kstar0"{name=T, below}]\arrow[uur, dash, "\kstar0\kstar"] && 0\kstar0 \\
      \arrow[rightarrow, from=F, to=T, phantom, "\kstar\kstar\kstar" description]
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\begin{definition}[\CSet]
  We define the category of semi-cubical sets as the functor category:
  \begin{equation*}
    \CSet := \Set^{\op{\Cube}}
  \end{equation*}
\end{definition}

\begin{definition}[$\Cube^n$]
  The standard semi-cube $\Cube^n$ is defined as the Yoneda embedding of $n \in \Obj(\Cube)$:
  \begin{align*}
    \Cube^n    & : \CSet                    \\
    \Cube^n(p) & := \Hom(p, n)              \\
    \Cube^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

\begin{example}[$\Cube^0$]
  $\Cube^0$ is $\Hom(0, 0)$, or the singleton set of the empty sequence:
  \begin{equation*}
    \begin{tikzcd}
      \{\}
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\Cube^1$]
  $\Cube^1$ consists of two points, given by $\Hom(0, 1)$ and a line, given by $\Hom(1, 1)$:
  \begin{equation*}
    \begin{tikzcd}
      L \arrow[r, "\kstar", dash] & R
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\Cube^2$]
  $\Cube^2$ consists of four points, given by $\Hom(0, 2)$, four lines connecting the four points, given by $\Hom(1, 2)$, and a filler, given by $\Hom(2, 2)$:
  \begin{equation*}
    \begin{tikzcd}
      LR \arrow[r, dash, "\kstar R"{name=F}] \arrow[d, dash, "L\kstar" left] & RR \arrow[d, dash, "R\kstar"] \\
      LL \arrow[r, dash, "\kstar L"{name=T, below}] & RL \\
      \arrow[rightarrow, from=F, to=T, phantom, "\kstar\kstar" description]
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\nu$-semi-shape category. The morphisms of the $\nu$-semi-shape category are sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$, so that the following holds.

@@

\section{Our setting}
We now explain how we recover the notion of sets in type theory, explain the difference between the fibered and indexed representations, and supply an example of indexed semi-cubical sets.

@@

A notion of sets can be recovered in each universe as $\U[m]$, denoting the subtype of $\Type_m$ for which paths are degenerated, what can be expressed by the property of Uniqueness of Identity Proofs (UIP). Technically, this is expressed as a structure equipping a domain $\Dom$ with the property $\UIP$:
\begin{align*}
  \Dom & : \Type_m                                  \\
  \UIP & : \Pi x y: \Dom.\, \Pi p q: x = y.\, p = q
\end{align*}

@@

Note that an equivalent definition of UIP is given by \textsf{UIP\_refl}:
\begin{align*}
  \textsf{UIP\_refl} & : \Pi x: \Dom.\, \Pi p: x = x.\, p = \textsf{eq\_refl } x
\end{align*}

@@

The latter correspondence can be expressed in Homotopy Type Theory, for $T: \Type$, as an equivalence between $\Sigma S: \Type\;.\;S \rightarrow T$ and $T \rightarrow \Type$, which can be proved using univalence. This correspondence also holds on \U, for $T: \U$:
\begin{equation*}
  \Sigma S: \U\;.\;S\rightarrow T ~\simeq~ T \rightarrow \U
\end{equation*}

\subsection{Example: indexed semi-cubical sets}
Let us now come back to semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0 : \U & X_1 : \U \arrow[l, "\delta^L" description, shift left=2] \arrow[l, "\delta^R" description, shift right=2] & X_2 : \U \arrow[l, "\delta^{L\kstar}" description, shift left=6] \arrow[l, "\delta^{R\kstar}" description, shift left=2] \arrow[l, "\delta^{\kstar R}" description, shift right=2] \arrow[l, "\delta^{\kstar L}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions. Based on these coherence conditions, the type of $X_2$ happens to have the form $\Pi a b c d, X_1 (a)(b) \rightarrow X_1 (a)(c) \rightarrow X_1 (b)(d) \rightarrow X_1 (c)(d) \rightarrow \U$, recovering the square structure of $X_2$. In summary,
\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

@@

\subsection{Relating the indexed representation to parametricity}
The process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$ in the last section, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. The parametricity which we consider interprets a type $A$ by a graph $A_\kstar$ over this type, and a term $t: A$ as an edge in $A_\kstar(t)(t)$. In particular, $\U$ is interpreted as the graph ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \rightarrow A_R \rightarrow \U$ of graphs over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
   & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq                                                      \\
   & \Pi a_L: A.\, \Pi a_R: A.\, \Pi a_\kstar: A_\kstar(a_L)(a_R).\, B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

In particular, for $X: \U$, applying the parametricity translation is about canonically associating to $X$ an inhabitant $X_\kstar$ of ${\U}_\kstar(X)(X)$ i.e. of $X \rightarrow X \rightarrow \U$. In turn, applying the parametricity translation to $X_\kstar: X \rightarrow X \rightarrow \U$ is about canonically associating to $X_\kstar$ an inhabitant $X_{\kstar\kstar}$ of $(X \rightarrow X \rightarrow \U)_\kstar(X_\kstar)(X_\kstar)$ i.e. of:
\begin{align*}
  \Pi x_{LL}: X.\, \Pi x_{LR}: X.\, X_\kstar(x_{LL})(x_{LR})             \\
  \rightarrow \Pi x_{RL}: X.\, \Pi x_{RR}: X.\, X_\kstar(x_{RL})(x_{RR}) \\ \rightarrow X_\kstar(x_{LL}, x_{RL}) \rightarrow X_\kstar(x_{LR})(x_{RR})  \rightarrow \U
\end{align*}
which hints us at how the sequence $X_0$, $X_1$, $X_2$ can be seen as a sequence of inhabitants of the iteration of binary parametricity applied to an initial $X: \U$:
\begin{equation*}
  \begin{array}{llllr}
    X_0 & \defeq & X                & : & \U                                               \\
    X_1 & \defeq & X_\kstar         & : & {\U}_\kstar(X)(X)                                \\
    X_2 & \defeq & X_{\kstar\kstar} & : & ({\U}_\kstar(X)(X))_{\kstar}(X_\kstar)(X_\kstar) \\
    \ldots
  \end{array}
\end{equation*}

@@

\subsection{Intuition for our formal construction}
To assign types to $X_0, X_1, X_2, \ldots$ in the above indexed representation, we give a recursive definition relying on building blocks, which we call $\framep$, $\layer$, and $\painting$. A $\framep$ is a boundary of a standard form (simplex, cube, etc.), which we decompose into $\layer$, and a $\painting$ corresponds to a filled $\framep$. Some $\framep$ are full and we call them $\fullframe$.

@@

We illustrated here the cubical case, that is $\nu = 2$, but, in general, a $\layer[n][p]$ is a product of $\nu$ $\painting[n - 1][p]$. A $\painting[n][0]$ is a $n$-dimensional object corresponding to a filled $\fullframe[n]$. More generally, a $\painting[n][p]$ is an $n$-dimensional object which has the form of a $\painting[n-p][0]$, thus of $(n-p)$-dimensional form, but shifted and living in dimensions $p$ to $n$. Such $\painting[n][p]$ fills a space framed by a partial $\framep[n][p]$ so that, together, they form a filled $\fullframe[n]$. For instance, in the picture, each of the two $\painting[2][0]$ of $\layer[3][0]$ is a filled blue square, each of the two $\painting[2][1]$ of $\layer[3][1]$ is the line shown in red, stretched into a partial square filling the partial frames made of respectively the left and right border of the blue square, and each of the two $\painting[2][2]$ of $\layer[3][2]$ is a point shown in green, stretched into a partial square filling the full frames made respectively of the upper and lower borders of the blue and red squares. A $\painting[n][p]$ adds to a $\framep[n][p]$ the layers needed to form a $\fullframe[n]$. They are added from dimension $n$ to dimension $p$ that is opposite to the order from $0$ to $p$ the $\framep[n][p]$ are built. A $\painting[n][p]$ complementing a given $\framep[n][p]$ fills the resulting $\fullframe[n]$. Thus, on the contrary of frames, it includes a filler, which lives in $X_n$ as shown below.
\begin{align*}
  \framep[n][p]   & \defeq \Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0]) \ldots) .\; \layer[n][p] \\
  \painting[n][p] & \defeq \Sigma l_p : \layer[n][p] . (\ldots (\Sigma l_n : \layer[n][n - 1] .\; X_n) \ldots)
\end{align*}

So far, we have not paid attention to the fact that we have a dependent type, shown as a $\Sigma$. Let us be more precise about this requirement. First, $\fullframe[n]$ depends on all $X_i$ up to $n - 1$. So, we need to package up $X_i$, for $i < n$, into a type, which we call $\X[][<n][]$. This allows us to give the type $\X[][<n][] \rightarrow \U$ to $\fullframe[n]$. Then, for $D: \X[][<n][]$, representing an initial prefix of $X_0, X_1, \ldots X_{n - 1}$, the indexed set $X_n$ has type $\fullframe[n](D) \rightarrow \U$. Thus, $\framep[n][p]$, $\layer[n][p]$ and $\painting[n][p]$ also depend on $D$. We can then refine the previous equation by showing the dependencies on $D$. In particular, $X_n$ is just $D.\tl$.
\begin{align*}
   & \framep[n][p](D) \defeq                                                                                     \\
   & \;\;\;\;\Sigma a_n : (\ldots (\Sigma \unitpoint : \unittype .\; \layer[n][0](D)) \ldots) .\,\layer[n][p](D) \\
   & \painting[n][p](D) \defeq                                                                                   \\
   & \;\;\;\;\Sigma l_p : \layer[n][p](D).\, (\ldots (\Sigma l_n : \layer[n][n - 1](D) .\, D.\tl) \ldots)
\end{align*}

@@

Let us recall that a layer is made of $\nu$ paintings. When $\nu = 2$, this amounts to:
\begin{align*}
  \layer[n][p](D)(d) & \defeq & \painting[n-1][p](D.\hd)(\restrf[n][p][L][p](d)) & \times \\
                     &        & \painting[n-1][p](D.\hd)(\restrf[n][p][R][p](d)) &
\end{align*}

@@

Given $c_\omega$ of type
\begin{align*}
  c_\omega & : \painting[n-1][p](D.\hd)(\restrf[n-1][p][\omega][q](d))
\end{align*}
$\restrl[n][p][\epsilon](d)(c_L, c_R)$ produces a layer, in which the $\omega$-component has the type
\begin{equation*}
  \painting[n-2][p](D.\hd.\hd)(\restrf[n-1][p][\epsilon][q](\restrf[n][p][\omega][p](d)))
\end{equation*}
while we expect a term of type
\begin{equation*}
  \painting[n-2][p](D.\hd.\hd)(\restrf[n-1][p][\omega][p](\restrf[n][p][\epsilon][q+1](d)))
\end{equation*}

@@

We now present the construction in \emph{extensional} type theory; i.e. in a type theory with the following reflection rule, where $=$ is propositional equality in some type and $\equiv$ is definitional equality~\cite{martinlof84}:
\begin{equation*}
  \seqr{}{\Gamma \vdash p : t = u}{\Gamma \vdash t \equiv u}
\end{equation*}

@@
